Quantifying Heterogeneity in SAR Imagery with the Rényi Entropy

### 2025-06-04 Text/Articles/RENYI_ENTROPY_SAR_2025/R1-Renyi-entropy-sar.qmd
R1 has been submitted

      This research, including its outputs, e.g., this manuscript, was executed in Quarto and is fully reproducible. We used RStudio version 2024.12.1+563, and R version 4.4.2. Data and code are available at: <https://github.com/rjaneth/renyi-entropy> 
    # bio: |
    #   Use `IEEEbiographynophoto` and the author name
    #   as the argument followed by the biography text.
    # note: "Template created June 23, 2023; revised `r format(Sys.Date(),format='%B %d, %Y')`."
abstract: |
  Quantifying surface roughness in synthetic aperture radar (SAR) data is critical for accurate geophysical interpretation and remote sensing applications. 
  We propose a test statistic based on a non-parametric estimation of Rényi entropy to characterize surface roughness from SAR intensity data. 
  The statistic is refined using bootstrap to improve its stability, size, and power. 
  This approach enhances roughness quantification by capturing scale-dependent variations and addressing data-driven uncertainty.
  Experimental results demonstrate the robustness of the proposed method in distinguishing roughness patterns.
  <!-- , offering a statistically rigorous tool for SAR-based terrain analysis. -->
  <!-- Synthetic aperture radar (SAR) systems have already been successfully used to solve remote sensing problems.  -->
  <!-- A disadvantage of SAR images is the presence of speckle, which is fully developed in homogeneous areas and gamma-distributed in these scenes.  -->
  <!-- In heterogeneous areas, the intensity values are $\mathcal{G}^0_I$-distributed.  -->
  <!-- In this way, the identification of roughness SAR areas (as opposed to homogeneous ones) is an important task.  -->
  <!-- In this work, we propose a family of hypothesis tests driven by an order parameter to identify roughness features in SAR intensity data using Rényi entropy.  -->
  <!-- In particular, we use a non-parametric estimator for the Rényi entropy and investigate some of its properties.  -->
  <!-- As a practical evaluation method, we develop $p$-value maps on which one can observe both (i) the heterogeneous evidence change per texture and (ii) the prediction of homogeneous and heterogeneous categories.  -->
  <!-- The results are in favor of the Rényi-based heterogeneity detector compared to the one based on Shannon entropy. -->
keywords: [Rényi entropy, Gamma distribution, heterogeneity, SAR, hypothesis tests, bootstrap]

#funding: 
 # statement: "The `quarto-ieee` "
pageheader:
  left: Journal XXX, Month Year
  right: #'D. Folio:  A Sample Article Using quarto-ieee'

@@

The $\mathcal{G}^0_I$ distribution effectively characterizes SAR intensity data because it captures different roughness levels. 
A limiting case is the Gamma distribution, associated with fully developed speckle in textureless regions.
However, selecting suitable models is challenging due to small window sizes in practical applications and difficulties in parameter estimation, motivating alternative statistical approaches.

@@

The main distributions considered for SAR intensity data are the $\Gamma_{\text{SAR}}$ distribution, which is suitable for fully developed speckle, and the $\mathcal{G}^0_I$ distribution, which can describe roughness&nbsp;[@Frery1997]. These distributions are characterized by the following probability density functions (pdfs):
\begin{equation}
	f_{\Gamma_{\text{SAR}}}\bigl(z;L, \mu \bigr) 
    = \frac{L^L}{\Gamma(L)\,\mu^L} z^{L-1} 
    \exp \biggl(-\frac{Lz}{\mu}\biggr)
    \mathbbm 1_{\mathbbm R_+}(z) \label{E:gamma1}
\end{equation}
and
\begin{multline}
    f_{\mathcal{G}^0_I}\bigl(z; \mu, \alpha, L \bigr) 
    = \frac{L^L\,\Gamma(L-\alpha)}
    {\bigl[-\mu(\alpha+1)\bigr]^{\alpha} \Gamma(-\alpha)\,\Gamma(L)}\\
    \frac{z^{L-1}}
    {\bigl[-\mu(\alpha+1)+Lz\bigr]^{L-\alpha}}
    \mathbbm 1_{\mathbbm R_+}(z), \label{E:gi01}
\end{multline}
where $\mu > 0$ is the mean,
$\alpha < 0$ measures the roughness, $L \geq 1$ is the number of
looks, $\Gamma(\cdot)$ is the gamma function, and
$\mathbbm 1_{A}(z)$ is the indicator function of the set $A$. 
The $\Gamma_{\text{SAR}}$  law is a particular case of the $\mathcal{G}^0_I$ distribution&nbsp;[@Frery1997]: for $\mu$ fixed,
$$
f_{\mathcal{G}^0_I}\big(z; \mu, \alpha, L\big)
\longrightarrow 
f_{\Gamma_{\text{SAR}}}(z;L, \mu) \text{ when } \alpha\to-\infty.
$$

@@

where the excess term $\Delta_\alpha$ depends on the roughness parameter.
When $\alpha\!\to\!-\infty$ (fully developed speckle), this excess tends to zero,
%<!-- , so the heterogeneous model collapses to the homogeneous one in Equation \eqref{eq-HGammaSAR}.  -->
and detecting heterogeneity is equivalent to deciding whether $\Delta_\alpha$ is significantly different from zero.
<!-- Note that the Rényi entropy of $\mathcal{G}^0_I$ can be expressed in terms of that of $\Gamma_{\mathrm{SAR}}$, with additional terms that depend on the roughness parameter $\alpha$. As $\alpha \to -\infty$, these terms tend to zero, reducing&nbsp;\eqref{eq-HGI0} to&nbsp;\eqref{eq-HGammaSAR}. -->
<!-- This work's core idea is to check if this excess entropy is significantly different from zero. -->
Fig.&nbsp;\ref{fig-convergence} illustrates this behaviour: as $\alpha$ decreases the colored curves of $H_\lambda(\mathcal{G}^0_I)$ converge towards the solid black curve of $H_\lambda(\Gamma_{\mathrm{SAR}})$.
<!-- , confirming the limiting behavior. -->

@@

  scale_color_manual(values = pal_jama()(7)[2:5], labels = alpha_labels) +
 # scale_color_manual(values = brewer.pal(4, "Dark2"), labels = alpha_labels) +
  # Labels and titles
  labs(color = "Roughness", 
       x = expression(mu), 
       y = "Rényi Entropy", 
       linetype = NULL) +

@@

The output consists of $p$-value maps and binary maps at a $0.05$ significance level. 
The $p$-value maps use a color gradient (Figs.&nbsp;\ref{fig:london-renyi},&nbsp;\ref{fig:london-shann},&nbsp;\ref{fig:munich-renyi},&nbsp;\ref{fig:munich-shann},&nbsp;\ref{fig:dublin-renyi},&nbsp;\ref{fig:dublin-shann}), where darker regions indicate areas with higher roughness, and lighter regions denote less textured surfaces. 
The binary maps (Figs.&nbsp;\ref{fig:london-005-renyi},&nbsp;\ref{fig:london-005-shann},&nbsp;\ref{fig:munich-005-renyi},&nbsp;\ref{fig:munich-005-shann},&nbsp;\ref{fig:dublin-005-renyi},&nbsp;\ref{fig:dublin-005-shann}) display heterogeneous detections in black ($p<0.05$) and homogeneous areas in white ($p \geq 0.05$).
Visually, the Rényi test exhibits greater sensitivity in detecting textural variations due to the flexibility provided by the parameter $\lambda$.

>>>>

      This research, including its outputs, e.g., this manuscript, was executed in Quarto and is fully reproducible. We used RStudio version 2024.12.1+563, and R version 4.4.2. Data and code are available at: <https://github.com/rjaneth/renyi-entropy> 
    # bio: |
    #   Use `IEEEbiographynophoto` and the author name
    #   as the argument followed by the biography text.
    # note: "Template created June 23, 2023; revised `r format(Sys.Date(),format='%B %d, %Y')`."
abstract: |
  Quantifying heterogeneity in synthetic aperture radar (SAR) data is critical for accurate geophysical interpretation and remote sensing applications. 
  We propose a test statistic based on a non-parametric estimation of Rényi entropy to characterize return heterogeneity from SAR intensity data. 
  The statistic is refined using bootstrap to improve its stability, size, and power. 
  This approach enhances heterogeneity quantification by capturing scale-dependent variations and addressing data-driven uncertainty.
  Experimental results establish the robustness of the proposed method in distinguishing heterogeneity patterns.
  <!-- , offering a statistically rigorous tool for SAR-based terrain analysis. -->
  <!-- Synthetic aperture radar (SAR) systems have already been successfully used to solve remote sensing problems.  -->
  <!-- A disadvantage of SAR images is the presence of speckle, which is fully developed in homogeneous areas and gamma-distributed in these scenes.  -->
  <!-- In heterogeneous areas, the intensity values are $\mathcal{G}^0_I$-distributed.  -->
  <!-- In this way, the identification of roughness SAR areas (as opposed to homogeneous ones) is an important task.  -->
  <!-- In this work, we propose a family of hypothesis tests driven by an order parameter to identify roughness features in SAR intensity data using Rényi entropy.  -->
  <!-- In particular, we use a non-parametric estimator for the Rényi entropy and investigate some of its properties.  -->
  <!-- As a practical evaluation method, we develop $p$-value maps on which one can observe both (i) the heterogeneous evidence change per texture and (ii) the prediction of homogeneous and heterogeneous categories.  -->
  <!-- The results are in favor of the Rényi-based heterogeneity detector compared to the one based on Shannon entropy. -->
keywords: [Rényi entropy, Gamma distribution, heterogeneity, SAR, hypothesis tests, bootstrap]

#funding: 
 # statement: "The `quarto-ieee` "
pageheader:
  left: IEEE Geoscience and Remote Sensing Letters, Month Year
  right: #'D. Folio:  A Sample Article Using quarto-ieee'

@@

The $\mathcal{G}^0_I$ distribution effectively characterizes SAR intensity data because it captures different heterogeneity levels. 
A limiting case is the Gamma distribution, associated with fully developed speckle in textureless regions.
However, selecting suitable models is challenging due to small window sizes in practical applications and difficulties in parameter estimation, motivating alternative statistical approaches.

@@

The main distributions considered for SAR intensity data are the $\Gamma_{\text{SAR}}$ distribution, which is suitable for fully developed speckle, and the $\mathcal{G}^0_I$ distribution, which can describe varying levels of heterogeneity&nbsp;[@Frery1997]. These distributions are characterized by the following probability density functions (pdfs):
\begin{equation}
	f_{\Gamma_{\text{SAR}}}\bigl(z;L, \mu \bigr) 
    = \frac{L^L}{\Gamma(L)\,\mu^L} z^{L-1} 
    \exp \biggl(-\frac{Lz}{\mu}\biggr)
    \mathbbm 1_{\mathbbm R_+}(z) \label{E:gamma1}
\end{equation}
and
\begin{multline}
    f_{\mathcal{G}^0_I}\bigl(z; \mu, \alpha, L \bigr) 
    = \frac{L^L\,\Gamma(L-\alpha)}
    {\bigl[-\mu(\alpha+1)\bigr]^{\alpha} \Gamma(-\alpha)\,\Gamma(L)}\\
    \frac{z^{L-1}}
    {\bigl[-\mu(\alpha+1)+Lz\bigr]^{L-\alpha}}
    \mathbbm 1_{\mathbbm R_+}(z), \label{E:gi01}
\end{multline}
where $\mu > 0$ is the mean,
$\alpha < 0$ measures the heterogeneity, $L \geq 1$ is the number of
looks, $\Gamma(\cdot)$ is the gamma function, and
$\mathbbm 1_{A}(z)$ is the indicator function of the set $A$. 
The $\Gamma_{\text{SAR}}$  law is a particular case of the $\mathcal{G}^0_I$ distribution&nbsp;[@Frery1997]: for $\mu$ fixed,
$$
f_{\mathcal{G}^0_I}\big(z; \mu, \alpha, L\big)
\longrightarrow 
f_{\Gamma_{\text{SAR}}}(z;L, \mu) \text{ when } \alpha\to-\infty.
$$

@@

where the excess term $\Delta_\alpha$ depends on the heterogeneity parameter.
When $\alpha\!\to\!-\infty$ (fully developed speckle), this excess tends to zero,
%<!-- , so the heterogeneous model collapses to the homogeneous one in Equation \eqref{eq-HGammaSAR}.  -->
and detecting heterogeneity is equivalent to deciding whether $\Delta_\alpha$ is significantly different from zero.
<!-- Note that the Rényi entropy of $\mathcal{G}^0_I$ can be expressed in terms of that of $\Gamma_{\mathrm{SAR}}$, with additional terms that depend on the roughness parameter $\alpha$. As $\alpha \to -\infty$, these terms tend to zero, reducing&nbsp;\eqref{eq-HGI0} to&nbsp;\eqref{eq-HGammaSAR}. -->
<!-- This work's core idea is to check if this excess entropy is significantly different from zero. -->
Fig.&nbsp;\ref{fig-convergence} illustrates this behaviour: as $\alpha$ decreases the colored curves of $H_\lambda(\mathcal{G}^0_I)$ converge towards the solid black curve of $H_\lambda(\Gamma_{\mathrm{SAR}})$.
<!-- , confirming the limiting behavior. -->

@@

  scale_color_manual(values = pal_jama()(7)[2:5], labels = alpha_labels) +
 # scale_color_manual(values = brewer.pal(4, "Dark2"), labels = alpha_labels) +
  # Labels and titles
  labs(color = "Heterogeneity", 
       x = expression(mu), 
       y = "Rényi Entropy", 
       linetype = NULL) +

@@

The output consists of $p$-value maps and binary maps at a $0.05$ significance level. 
The $p$-value maps use a color gradient (Figs.&nbsp;\ref{fig:london-renyi},&nbsp;\ref{fig:london-shann},&nbsp;\ref{fig:munich-renyi},&nbsp;\ref{fig:munich-shann},&nbsp;\ref{fig:dublin-renyi},&nbsp;\ref{fig:dublin-shann}), where darker regions indicate areas with higher heterogeneity, and lighter regions denote less textured surfaces. 
The binary maps (Figs.&nbsp;\ref{fig:london-005-renyi},&nbsp;\ref{fig:london-005-shann},&nbsp;\ref{fig:munich-005-renyi},&nbsp;\ref{fig:munich-005-shann},&nbsp;\ref{fig:dublin-005-renyi},&nbsp;\ref{fig:dublin-005-shann}) display heterogeneous detections in black ($p<0.05$) and homogeneous areas in white ($p \geq 0.05$).
Visually, the Rényi test exhibits greater sensitivity in detecting textural variations due to the flexibility provided by the parameter $\lambda$.

### 2025-06-04 Text/Articles/RENYI_ENTROPY_SAR_2025/Submissions/R1/Responses/R1-CoverLetter.tex
R1 has been submitted

\begin{letter}{Dr.\ Ronny Hänsch\\
	Editor-in-Chief\\
	IEEE Geoscience and Remote Sensing Letters}
\date{1 April 2025}

@@

We want to inform that we submitted the manuscript GRSL-00722-2025 (``Quantifying Roughness in SAR Imagery with the Rényi Entropy'') to the MIGARS Special Stream because it is a significantly extended version of ``Identifying departures from the fully developed speckle hypothesis in intensity SAR data with non-parametric estimation of the entropy'' (DOI \url{10.1109/MIGARS61408.2024.10544448}), presented at MIGARS~2024 and already published in IEEExplore.

>>>>

\begin{letter}{Dr.\ Ronny Hänsch\\
	Editor-in-Chief\\
	IEEE Geoscience and Remote Sensing Letters}
\date{4 June 2025}

@@

We appreciate the efforts in finding a suitable Associate Editor and reviewer for our manuscript.

We want to clarify that we focused our revision on the comments provided by the Associate Editor and Reviewer \#1.
We didn't find the Reviewer \#2 comments constructive or specific.
Nevertheless, what this last reviewer criticized was also improved thanks to our handling of the other comments.
